Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-02 manuscript.qmd

To be able to create models for lymphatic tumor progression that includes all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consists of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.

@@

1. More advanced T-category should lead to higher risk for nodal disease. One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
2. The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease. This should come in addition to what can be inferred about this time from T-category alone.
3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.

@@

$$
\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
$$ {#eq-single-patient-llh}

@@

For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube. Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated. For short chains, this estimate is not trustworthy, but stabilizes for longer chains. We therefore considered a sampling to be converged when two criteria where met:

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only crosses a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->. When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.

>>>>

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.

@@

1. More advanced T-category should lead to higher risk for nodal disease. One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
2. The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease. This should come in addition to what can be inferred about the time from T-category alone.
3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.

@@

$$
\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
$$ {#eq-single-patient-llh}

@@

For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube. Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated. For short chains, this estimate is not trustworthy, but stabilizes for longer chains. We therefore considered a sampling to be converged when two criteria were met:

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->. When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.
